---
title: "Project Work"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Project Introduction

The goal of this project is to create a set of reusable R functions that can query, process, summarize, and visualize data from the Public Use Microdata Sample (PUMS) Census API. The PUMS data provide person-level survey information collected through the American Community Survey, allowing researchers to investigate demographic, economic, educational, and geographic patterns across the United States. Rather than manually constructing API requests and cleaning the returned data each time, this project focuses on developing functions that automate the entire workflow.

The first step in this project is to establish a reliable process for obtaining data from the Census PUMS API. Before creating more advanced functions, it is important to verify that a request can be successfully sent to the API and that the returned data can be imported into R.

```{r}
library(httr)
library(jsonlite)
library(tibble)
library(tidyverse)

# Census API URL (example)
url <- "https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&for=division:*&SCHL=24&key=eb86df934ff00b4bbaa27d768425acced8a927e0"

url_response <- GET(url)
parsed <- fromJSON(rawToChar(url_response$content))

# retrieve column names
col_names <- parsed[1, ]
data <- parsed[-1, ]

# turn the data into a tibble and apply the column names
census_data <- as_tibble(data)
names(census_data) <- col_names

census_data
```

Now, lets write a helper function to take what is returned by GET() and turn it into a nice tibble. This will automate what we did above.

```{r}
helper_fun <- function(response) {
  parsed <- fromJSON(rawToChar(response$content))
  col_names <- parsed[1, ]
  data <- parsed[-1, ]
  df <- as_tibble(data)
  names(df) <- col_names
  df
}
```

```{r}
# test the function to see if it works
helper_fun(url_response)

# test it again using a different url
url_response_2 <- GET("https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=eb86df934ff00b4bbaa27d768425acced8a927e0")
helper_fun(url_response_2)

# everything seems to work correctly!
```

In the previous step, we established a working workflow for retrieving data from the Census API and converting the raw JSON response into a clean tibble. While this confirms that the API connection and parsing process works, a fixed URL approach is not sufficient for exploratory analysis.

To make the data pipeline more flexible and reusable, the next step is to build a function that allows users to specify key components of the API request. This includes the survey year, selected variables, and geographic scope of the data being returned.

Before we do that, we first need to create the numeric_time function to clean and standardize the `JWAP` and `JWDP` variables. These variables are not stored as traditional numeric values. Instead, they are coded as time intervals that represent ranges of minutes or hours during which an event (such as arrival or departure from work) occurred. Because these values are interval-based rather than numeric, they cannot be directly used in calculations such as averages or other summary statistics.

The purpose of the `numeric_time` function is to convert these coded values into a usable numeric format. It begins by coercing the input into a numeric type to ensure consistency across observations. It then replaces any values less than or equal to zero with NA, since these represent cases where the variable is not applicable.

```{r}
numeric_time <- function(x) {
  x <- as.numeric(x)
  x[x <= 0] <- NA
  return(x)
}
```


```{r}
census_api_fun <- function(year = 2022,
                           numeric_vars = c("AGEP", "PWGTP"),
                           categorical_vars = c("SEX"),
                           geography = "State",
                           geo_subset = "37",
                           key = "eb86df934ff00b4bbaa27d768425acced8a927e0") {
  
  # check year
  if (!is.numeric(year) || year < 2010 || year > 2022) {
    stop("Year must be between 2010 and 2022.")
  }
  
  # allowed variables 
  allowed_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  
  numeric_vars <- unique(c("PWGTP", numeric_vars))
  
  if (!all(numeric_vars %in% allowed_numeric | numeric_vars == "PWGTP")) {
    stop("Invalid numeric variable selected.")
  }
  
  if (!all(categorical_vars %in% allowed_categorical)) {
    stop("Invalid categorical variable selected.")
  }

  allowed_geo <- c("Region", "Division", "State")
  
  if (!(geography %in% allowed_geo)) {
    stop("Geography must be Region, Division, or State.")
  }
  
  geo_part <- switch(geography,
                     "State" = paste0("state:", geo_subset),
                     "Division" = "division:*",
                     "Region" = "region:*")
  
  # build URL
  vars <- paste(c(numeric_vars, categorical_vars), collapse = ",")
  
  url <- paste0("https://api.census.gov/data/",
                year,
                "/acs/acs1/pums?get=",
                vars,
                "&for=", geo_part,
                "&key=", key)
  
  response <- GET(url)
  raw_text <- rawToChar(response$content)
  
  if (response$status_code != 200 || grepl("<html", raw_text)) {
    stop(paste("API request failed for year", year, "\nURL:", url))
  }
  
  parsed <- fromJSON(raw_text)
  
  # make it a tibble
  col_names <- parsed[1, ]
  data <- parsed[-1, ]
  
  df <- as_tibble(data)
  names(df) <- col_names
  
  for (v in numeric_vars) {
    if (v %in% c("JWAP", "JWDP")) {
      df[[v]] <- numeric_time(df[[v]])
    } else {
      df[[v]] <- as.numeric(df[[v]])
    }
  }
  
  for (v in categorical_vars) {
    df[[v]] <- as.factor(df[[v]])
  }
  
  class(df) <- c("census", class(df))
  
  return(df)
}
```

Lets test our function out. We'll test it once without the optional `geography` argument, once with a `year` outside the accepted range, and once with one of the time interval variables (`JWDP` and `JWAP`).

```{r}
# this should work, leave out the optional geography argument
census_api_fun(year = 2021, numeric_vars = c('AGEP', 'GASP'), categorical_vars = c('SEX', 'FER'))

# this should not work, year is out of bounds
# I've commented this out to render the document
#census_api_fun(year = 2009, numeric_vars = c('AGEP', 'GASP'), categorical_vars = c('SEX', #'FER'), geography = 'State')

# Lets see how our time intervals do
census_api_fun(year = 2015, numeric_vars = c('AGEP', 'JWDP'), categorical_vars = c('SEX', 'FER'), geography = 'Region')
```


To extend the flexibility of the data pipeline, a final function was developed to allow users to retrieve Census PUMS data across multiple survey years. While the single-year function provides a complete workflow for querying and cleaning data from one specific year, many analyses require examining changes over time. Rather than requiring the user to manually call the function multiple times and combine results, this process is automated through a wrapper function.

The multi-year function accepts a vector of years as input along with all of the same arguments used in the single-year function, including numeric variables, categorical variables, geographic level, and geographic subset. This ensures consistency in data processing across all years. As part of the function design, 2020 is explicitly removed from the input years, since the Census Bureau advises against using that year’s PUMS data due to disruptions caused by the COVID-19 pandemic.

The function operates by iterating over each remaining year and calling the single-year census API function for each iteration. After each dataset is retrieved and cleaned, a new variable is added to indicate the survey year. This step is important because once the datasets are combined, the year variable serves as the key identifier for comparing observations across time.

```{r}
census_api_multi_year <- function(years,
                                  numeric_vars = c("AGEP", "PWGTP"),
                                  categorical_vars = c("SEX"),
                                  geography = "State",
                                  geo_subset = "37",
                                  key = "eb86df934ff00b4bbaa27d768425acced8a927e0") {
  
  # remove 2020
  years <- years[years != 2020]
  
  if (length(years) == 0) {
    stop("No valid years provided (2020 is not allowed).")
  }
  
  # loop over years
  results <- lapply(years, function(y) {
    
    df <- census_api_fun(year = y,
                         numeric_vars = numeric_vars,
                         categorical_vars = categorical_vars,
                         geography = geography,
                         geo_subset = geo_subset,
                         key = key)

    df$year <- y
    df
  })
  
  # combine results
  final_df <- bind_rows(results)
  
  if (nrow(final_df) == 0) {
    stop("All API requests failed. Check variables, geography, or API limits.")
  }
  
  return(final_df)
}
```

Lets test our function out.

```{r}
census_api_multi_year(years = c(2015, 2017, 2021),
                      numeric_vars = c("AGEP", "JWDP"),
                      categorical_vars = c("SEX", "FER"),
                      geography = "State"
)
```

Great! Everything seems to be working as it should. 

Usually, our next step would be to process the data in our tibble appropriately. The `PWGTP` variable actually represents the number of (people) observations there should be for a particular row. We won’t do this, but we could create a new tibble with each row replicated `PWGTP` times. (This would create very large data frames in some instances and processing would be slow!). 

Up to this point, we have built a flexible pipeline for querying and combining Census PUMS data. While this allows us to retrieve structured data across different years and geographic levels, we still need a way to efficiently summarize the resulting datasets.

Rather than writing separate summary functions for each dataset type, we can take advantage of object-oriented programming in R, specifically S3 method dispatch. In R, generic functions such as `plot()` or `summary()` determine which method to use based on the class of the input object. For example, `plot()` behaves differently for data frames, functions, and time series objects because R internally calls `UseMethod("plot")` and then dispatches the appropriate method based on the object’s class.

We can use this same idea to create a custom summary method for our Census data. By assigning a new class called "census" to our tibble, we allow R to recognize our dataset as a specialized object type. This enables us to define a custom method, `summary.census()`, that automatically computes meaningful statistics tailored to the structure of our data.

Specifically, our summary function will:

- Compute weighted means and weighted standard deviations for numeric variables (excluding `PWGTP`, which is used as the weight itself)

- Compute weighted counts for categorical variables

- Allow the user to optionally select which variables to summarize, while defaulting to all valid variables in the dataset

This approach ensures that summaries properly reflect the survey design by incorporating the person-level weights (`PWGTP`) rather than treating each observation equally.

```{r}
summary.census <- function(object,
                           numeric_vars = NULL,
                           categorical_vars = NULL) {
  
  if (is.null(numeric_vars)) {
    numeric_vars <- names(object)[sapply(object, is.numeric)]
    numeric_vars <- setdiff(numeric_vars, c("PWGTP"))
  }
  
  if (is.null(categorical_vars)) {
    categorical_vars <- names(object)[sapply(object, is.factor)]
  }
  
  w <- object$PWGTP
  
  # numeric summaries
  numeric_summary <- lapply(numeric_vars, function(v) {
    
    x <- object[[v]]
    
    mean_w <- sum(x * w, na.rm = TRUE) / sum(w, na.rm = TRUE)
    
    sd_w <- sqrt(
      sum((x^2) * w, na.rm = TRUE) / sum(w, na.rm = TRUE) - mean_w^2
    )
    
    list(
      variable = v,
      weighted_mean = mean_w,
      weighted_sd = sd_w
    )
  })
  
  # categorical summaries (weighted counts)
  categorical_summary <- lapply(categorical_vars, function(v) {
    
    x <- object[[v]]
    counts <- tapply(w, x, sum, na.rm = TRUE)
    list(variable = v, weighted_counts = counts)
    
  })
  
  # return the output
  result <- list(numeric_summary = numeric_summary, categorical_summary = categorical_summary)
  
  class(result) <- "summary.census"
  return(result)
}
```


Lets test out this function by running `summary(census_api_fun(year = 2021))`.

```{r}
data <- census_api_fun(year = 2021)
summary(data)
```

Building on the custom `summary()` method, we can further extend the functionality of our "census" class by defining a custom plotting method. This allows us to use the generic `plot()` function while automatically producing a visualization tailored specifically to census survey data.

In this case, we define a `plot.census()` method that creates a weighted boxplot comparing a numeric variable across the levels of a categorical variable. The inclusion of the survey weight variable `PWGTP` in the aesthetic mapping is important because each row in the dataset does not represent a single individual equally. Instead, the weights adjust each observation so that the resulting plot reflects the true population structure.

To make the function flexible, the user specifies the categorical and numeric variables as column names. These are then accessed directly from the tibble using double brackets like so: (`[[ ]]`). This allows for clean and reliable selection of variables.

For this function, we'll need both the `tidyverse` package, so make sure it is installed on your computer. You can install this packages by running `install.packages("tidyverse")`.

```{r}
library(tidyverse)
plot.census <- function(x, cat_var, num_var) {
  
  if (missing(cat_var) || missing(num_var)) {
    stop("You must specify one categorical variable and one numeric variable.")
  }
  
  if (!(cat_var %in% names(x)) || !(num_var %in% names(x))) {
    stop("One or both variables are not in the dataset.")
  }
  
  # extract columns 
  x_cat <- x[[cat_var]]
  x_num <- x[[num_var]]
  
  # build the boxplot
  ggplot(x, aes(x = x_cat, y = x_num, weight = PWGTP)) +
    geom_boxplot() +
    labs(
      x = cat_var,
      y = num_var,
      title = paste("Weighted distribution of", num_var, "by", cat_var)
    )
}
```


Let's test out our function.

```{r}
plot.census(x = data, cat_var = "SEX", num_var = "AGEP")
```


# Investigating Age Differences Across Educational Attainment in North Carolina

Now that the Census API function, summary method, and plotting tools have been developed, we can use them to explore a substantive question in the data. One interesting relationship to examine is how age varies across different levels of educational attainment within North Carolina.

In this analysis, we use `AGEP` (age) as the numeric variable and `SCHL` (educational attainment) as the categorical grouping variable. This allows us to assess how the age distribution differs across the most common education categories in the population.

We begin by pulling the relevant variables from the Census API using our custom function:

```{r}
data <- census_api_fun(year = 2021,
                       numeric_vars = c("AGEP"),
                       categorical_vars = c("SCHL"),
                       geography = "State")
```


Because `SCHL` contains many possible categories, many of which occur infrequently, we reduce complexity by keeping only the four most common education categories. This step ensures that the resulting visualization remains interpretable. 


```{r}
data <- data[data$SCHL %in% names(sort(table(data$SCHL), decreasing = TRUE)[1:4]), ]
```

Next, we apply the custom summary method for objects of class census to obtain weighted descriptive statistics for age and counts of education categories:

```{r}
summary.census(data)
```

We then visualize the relationship using the custom weighted boxplot function:

```{r}
plot.census(data, cat_var = "SCHL", num_var = "AGEP")
```

This plot displays the distribution of age across the most common educational attainment categories while incorporating survey weights. This ensures that each observation contributes proportionally to the population it represents.












```{r}
data <- census_api_fun(
  year = 2021,
  numeric_vars = c("AGEP"),
  categorical_vars = c("SCHL"),
  geography = "State",
  geo_subset = "37"
)

data <- data[data$SCHL %in% names(sort(table(data$SCHL), decreasing = TRUE)[1:4]), ]
summary.census(data)

plot.census(data, cat_var = "SCHL", num_var = "AGEP")
```


