---
title: "Project Work"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Project Introduction

The goal of this project is to create a set of reusable R functions that can query, process, summarize, and visualize data from the Public Use Microdata Sample (PUMS) Census API. The PUMS data provide person-level survey information collected through the American Community Survey, allowing researchers to investigate demographic, economic, educational, and geographic patterns across the United States. Rather than manually constructing API requests and cleaning the returned data each time, this project focuses on developing functions that automate the entire workflow.

# Retrieving Data from the Census API

The first step in this project is to establish a reliable process for obtaining data from the Census PUMS API. Before creating more advanced functions, it is important to verify that a request can be successfully sent to the API and that the returned data can be imported into R.

```{r}
library(httr)
library(jsonlite)
library(tibble)

# Census API URL (fixed example)
url <- "https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&for=division:*&SCHL=24&key=eb86df934ff00b4bbaa27d768425acced8a927e0"

# send request
url_response <- GET(url)

# parse JSON content
parsed <- fromJSON(rawToChar(url_response$content))

# retrieve column names
col_names <- parsed[1, ]
data <- parsed[-1, ]

# turn the data into a tibble and apply the column names
census_data <- as_tibble(data)
names(census_data) <- col_names

census_data
```

Now, lets write a helper function to take what is returned by GET() and turn it into a nice tibble. This will automate what we did above.

```{r}
helper_fun <- function(response) {
  parsed <- fromJSON(rawToChar(response$content))
  col_names <- parsed[1, ]
  data <- parsed[-1, ]
  df <- as_tibble(data)
  names(df) <- col_names
  df
}

# test the function to see if it works
helper_fun(url_response)

# test it again using a different url
url_response_2 <- GET("https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=eb86df934ff00b4bbaa27d768425acced8a927e0")
helper_fun(url_response_2)

# everything seems to work correctly!
```

# Building a Flexible Census API Query Function

In the previous step, we established a working workflow for retrieving data from the Census API and converting the raw JSON response into a clean tibble. While this confirms that the API connection and parsing process works, a fixed URL approach is not sufficient for exploratory analysis.

To make the data pipeline more flexible and reusable, the next step is to build a function that allows users to dynamically specify key components of the API request. This includes the survey year, selected variables, and geographic scope of the data being returned.

Before writing this function, lets write a helper function that will turn time intervals into a numeric variable by taking the midpoint of the interval. 

```{r}
midpoint_time <- function(x) {

  x <- as.numeric(x)

  x[x <= 0] <- NA

  return(x)
}
```


```{r}
census_api_fun <- function(year = 2022,
                           numeric_vars = c("AGEP", "PWGTP"),
                           categorical_vars = c("SEX"),
                           geography = "State",
                           geo_subset = "37",
                           key = "eb86df934ff00b4bbaa27d768425acced8a927e0") {
  
  # ---------------------------
  # 1. VALIDATE YEAR
  # ---------------------------
  if (!is.numeric(year) || year < 2010 || year > 2022) {
    stop("Year must be between 2010 and 2022.")
  }
  
  # ---------------------------
  # 2. ALLOWED VARIABLES
  # ---------------------------
  allowed_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  
  numeric_vars <- unique(c("PWGTP", numeric_vars))
  
  if (!all(numeric_vars %in% allowed_numeric | numeric_vars == "PWGTP")) {
    stop("Invalid numeric variable selected.")
  }
  
  if (length(categorical_vars) < 1) {
    stop("You must select at least one categorical variable.")
  }
  
  if (!all(categorical_vars %in% allowed_categorical)) {
    stop("Invalid categorical variable selected.")
  }
  
  if (all(numeric_vars == "PWGTP")) {
    stop("Must include at least one numeric variable besides PWGTP.")
  }
  
  # ---------------------------
  # 3. GEOGRAPHY HANDLING (FIXED)
  # ---------------------------
  allowed_geo <- c("Region", "Division", "State")
  
  if (!(geography %in% allowed_geo)) {
    stop("Geography must be Region, Division, or State.")
  }
  
  if (geography == "State") {
    geo_part <- paste0("state:", geo_subset)
  } else if (geography == "Division") {
    geo_part <- "division:*"
  } else if (geography == "Region") {
    geo_part <- "region:*"
  }
  
  # ---------------------------
  # 4. BUILD VARIABLE STRING
  # ---------------------------
  vars <- paste(c(numeric_vars, categorical_vars), collapse = ",")
  
  # ---------------------------
  # 5. BUILD URL
  # ---------------------------
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums?get=",
    vars,
    "&for=", geo_part,
    "&key=", key
  )
  
  # ---------------------------
  # 6. GET DATA
  # ---------------------------
  response <- GET(url)
  parsed <- fromJSON(rawToChar(response$content))
  
  # ---------------------------
  # 7. TO TIBBLE
  # ---------------------------
  col_names <- parsed[1, ]
  data <- parsed[-1, ]
  
  df <- as_tibble(data)
  names(df) <- col_names
  
  # ---------------------------
  # 8. TYPE CONVERSION
  # ---------------------------
  
  for (v in numeric_vars) {
    if (v %in% c("JWAP", "JWDP")) {
      df[[v]] <- midpoint_time(df[[v]])
    } else {
      df[[v]] <- as.numeric(df[[v]])
    }
  }
  
  for (v in categorical_vars) {
    df[[v]] <- as.factor(df[[v]])
  }
  
  return(df)
}
```

Lets test our function out

```{r}
# this should work
census_api_fun(year = 2021, numeric_vars = c('AGEP', 'GASP'), categorical_vars = c('SEX', 'FER'), geography = 'State')

# this should not work, year is out of bounds
census_api_fun(year = 2009, numeric_vars = c('AGEP', 'GASP'), categorical_vars = c('SEX', 'FER'), geography = 'State')

# Lets see how our time intervals do
census_api_fun(year = 2015, numeric_vars = c('AGEP', 'JWDP'), categorical_vars = c('SEX', 'FER'), geography = 'Division')
```


